\pdftitle

### 2018-05-15 maxNt_v3.tex
V3

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-15T12:04:44+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$; \enquote{$X$} denote the
variable, and \enquote{$R$} its value. A particular sample of $n$ neurons
from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$ having
fixed binary values $(r_1, \dotsc, r_n)\defs\yr \in \set{0,1}^n$. We will
consider various averages of the population and the sample. For this
purpose we introduce a population-averaging operator $\sav{\dotv}$ defined
by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3 + \dotsb +
    X_{N-1} X_N),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{N}{3}^{-1} (X_1 X_2 X_3 +% X_1 X_2 X_4
    \dotsb + X_{N-2} X_{N-1}X_N),
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
and so on.
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

@@

In this state of knowledge, denoted $\yH$, we know that the neurons in the
population are biologically or functionally similar, for example in
morphology or the kind of input or output they receive or give. But we are
completely ignorant about the physical details of the individual neurons.
Our ignorance is therefore symmetric under permutations of neuron
identities. This ignorance leads us to assign a probability distribution
for the population state $\yX$ that is symmetric under permutations of
neuron identities; such a distribution is usually called \emph{finitely
  exchangeable} \cites{ericson1969}[\chap~1]{ghoshetal1997}. We stress that
this probability assignment is just an expression of the symmetry of our
\emph{ignorance} about the state of the population, not an expression of
some biologic or physical identity of the neurons.

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
The conditional probability in the last formula is a hypergeometric
distribution $\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of the formulae above is
in fact the same as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
Our initial symmetric knowledge should intuitively also apply to the
sample; indeed, the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

How is it possible that the very different states of knowledge $\yHp$ and
$\yHi$ lead to the same formulae above? Their difference appears as soon as
we make an experimental observation, say $X_2 = R_2\in\set{0,1}$ and update our
initial probabilities \eqref{eq:joint_plaus_N_homog}:
\begin{equation}
  \label{eq:updated_joint}
    \p(\yX = \yR \| X_2 = R_2, \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, X_2 = R_2, \yH)\,
  \p(\yXf=\yRf \| X_2 = R_2,\yH).
\end{equation}
The conditional probability and the probability for the average on the
right side will update in very different ways for $\yHp$ and $\yHi$. The
discussion of this update process in the two cases is unnecessary for the
purposes of this note and outside their scope; we hope to address it in
full in a future work.

The conditional probability $\ypp(\yrs,\yRf)$ relates the spaces of the
sample average $\yXf\in\set{0,\dotsc,N}$ and of the population average
$\yxs\in\set{0,\dotsc,n}$ in a special way. It is a coarsening projector of
any probability $p$ for $\yXf$ onto a marginal probability $p_*$ for
$\yxs$:
\begin{equation}
  \label{eq:projection_P}
  % \ypp\colon p(\yXf=\dotv) \to
  p_*(\yxs=\yrs)
  = \sum_{N\yRf=0}^N \ypp(\yrs\|\yRf) p(\yXf=\yRf).
\end{equation}
Conversely it also a pulls back expectations of functions $f$ of the sample
average $\yxs$ to expectations of functions $f^*$ of the population average
$\yXf$:
\begin{equation}
  \label{eq:pullback_P}
%  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n
%  f(\yrs)\ypp(\yrs\|\dotv),
  \begin{gathered}
    f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
    \\
   \expeb{f(\yxs)}= \expeb{f^*(\yXf)}
    =
    \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
    =
    \sum_{N\yRf=0}^N f^*(\yRf)\,\p(\yXf=\yRf \| \yH).
  \end{gathered}
\end{equation}

@@

Yet, the approximate equalities above express the fact that \emph{our
  uncertainty about the sample is representative of our uncertainty about
  the population and about other samples}, and vice versa. This fact comes
about for very different reasons in the states of knowledge $\yHp$ and
$\yHi$. In $\yHp$, because our sample is \emph{physically} representative
of the population and of other samples; we could write this as
$\yxs\approx\yXf$. In $\yHi$, because we are ignorant about sample and
population in similar symmetric ways; but this does \emph{not} imply that
$\yxs\approx\yXf$. New observations may in fact break this symmetry via
\eqn~\eqref{eq:updated_joint}.

Note that formulae~\eqref{eq:roughly_equal_nN} say more than the limits
$\p(\yxs=a) \to \p(\yXf=a)$ and $\expeb{f(\yxs)} \to \expeb{f(\yXf)}$, as
$n\to N$, do. These limits are trivially valid because the sample becomes
the full population as $n\to N$. In particular, these limits hold even in
cases where the conditional probability $\p(\yxs = \yrs \|\yXf=\yRf)$ is
not a fuzzy identity and our uncertainties about sample and about
population can differ wildly.

@@

% The projection~\eqref{eq:projection_P} and pull-back~\eqref{eq:pullback_P}
% have the property
% \begin{equation}
%   \label{eq:expectation_pullback}
%   \expeb{f(\yxs) \|\yH} = \expeb{f^*(\yXf)\|\yH}
%   \qquad\text{or}\qquad
%   \sum_{n\yrs=0}^n f(\yrs)\,p_*(\yxs=\yrs) = \sum_{N\yRf=0}^N f^*(\yRf)\,p(\yXf=\yRf).
% \end{equation}
For functions representing averaged products,
$f(\yxs) = \sav{\yx\dots\yx}\equiv\binom{n\yxs}{m}/\binom{n}{m}$,
formulae~\eqref{eq:pullback_P} have the useful form
\begin{equation}
  \label{eq:expe_products}
%  \label{eq:pullback_m_expectations}
%  \ypp^*\colon \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}
%  \mapsto
  \begin{gathered}
    \bigl( \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}\bigr)^* =
    \av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}},
    \\[2\jot]
    % \expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH} =
    % \expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
    \expe{\sav{\yx \dotsm \yx} \| \yH} =
    \expe{\av{\yX \dotsm \yX} \| \yH}
    % \\[2\jot]
    % {}
    = \binom{n}{m}^{-1}
    % \frac{(n-m)!}{n!}
    \sum_{\mathclap{n \yrs=0}}^n %m!\,
    \binom{n \yrs}{m}\, \p(\yxs=\yrs \| \yH)
    % \sum_{N \yRf=0}^N %m!\,
    % \binom{N-m}{N \yRf-m} \binom{N}{N \yRf}^{-1} \p(\yXf=\yRf \| \yH)
    = \binom{N}{m}^{-1}
    % \frac{(N-m)!}{N!}
    \sum_{\mathclap{N \yRf=0}}^N %m!\,
    \binom{N \yRf}{m}\, \p(\yXf=\yRf \| \yH).
  \end{gathered}
\end{equation}
The proof uses the expression for the $m$th factorial moment of the
hypergeometric distribution \cite{potts1953}. Thus, in the states of
knowledge $\yHp$ and $\yHi$ the averages of activity products \emph{are
  initially the same for the sample and for the full population}. Similar
relations can be found for the raw moments $\expe{\yxs^m}$ and
$\expe{\yXf^m}$, which can be written in terms of the product expectations
via \eqn~\eqref{eq:products_intermsof_average}.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-15T13:47:43+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$; the \enquote{$X$} denote the
variables, and \enquote{$R$} their values. A particular sample of $n$ neurons
from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$ having
fixed binary values $(r_1, \dotsc, r_n)\defs\yr \in \set{0,1}^n$. We will
consider various averages of the population and the sample. For this
purpose we introduce a population-averaging operator $\sav{\dotv}$ defined
by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3 + \dotsb +
    X_{N-1} X_N),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{N}{3}^{-1} (X_1 X_2 X_3 +% X_1 X_2 X_4
    \dotsb + X_{N-2} X_{N-1}X_N),
    \qquad\text{and so on.}
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

@@

In this state of knowledge, denoted $\yH$, we know that the neurons in the
population are biologically or functionally similar, for example in
morphology or the kind of input or output they receive or give. But we are
completely ignorant about the physical details of the individual neurons.
Our ignorance is therefore symmetric under permutations of neuron
identities. This ignorance is represented by a probability distribution
that is symmetric under permutations of neuron identities; such a
distribution is usually called \emph{finitely exchangeable}
\cites{ericson1969}[\chap~1]{ghoshetal1997}. We stress that this
probability assignment is just an expression of the symmetry of our
\emph{ignorance} about the state of the population, not an expression of
some biologic or physical identity of the neurons.

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
The conditional probability in the last formula is a hypergeometric
distribution $\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of the formulae above is
in fact the same as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
Our initial symmetric ignorance should intuitively also apply to the
sample; the probability for the state of the
sample~\eqref{eq:subpop_average} indeed satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

The conditional probability
$\p(\yxs = \yrs \|\yXf=\yRf, \yH) \equiv \ypp(\yrs,\yRf)$ relates the
spaces of the sample average $\yXf\in\set{0,\dotsc,N}$ and of the
population average $\yxs\in\set{0,\dotsc,n}$. It is a coarsening projector
of any probability for $\yXf$ onto a marginal probability for $\yxs$, as
\eqn~\eqref{eq:subpop_average} shows. For each function $f$ of the sample
average $\yxs$, the  hypergeometric distribution 
$\ypp(\yrs,\yRf)$ also allows us to construct a function $f^*$ of the
population average $\yXf$
\begin{equation}
  \label{eq:pullback_P}
    f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
%  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n
%  f(\yrs)\ypp(\yrs\|\dotv),
\end{equation}
that has the same expectation as $f$:
\begin{multline}
  \expeb{f^*(\yXf)}
  \equiv
  \sum_{N\yRf=0}^N f^*(\yRf)\,\p(\yXf=\yRf \| \yH)
  =
  \sum_{N\yRf=0}^N \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yRf)\,\p(\yXf=\yRf \| \yH)
  ={}\\
  \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
  \equiv  \expeb{f(\yxs)}
\end{multline}

@@

The approximate equalities above, on the other hand, express the fact that
\emph{our uncertainty about the sample is representative of our uncertainty
  about the population and about other samples}, and vice versa. These
approximate equalities are not generally valid for other probability
distributions for the population. They should be contrasted with the limits
$\p(\yxs=a) \to \p(\yXf=a)$ and $\expeb{f(\yxs)} \to \expeb{f(\yXf)}$, as
$n\to N$, do: these limits are trivially, universally valid because the
sample becomes the full population as $n\to N$. In particular, these limits
hold even in cases where the conditional probability
$\p(\yxs = \yrs \|\yXf=\yRf)$ is not a fuzzy identity, our uncertainties
about sample and about population can differ wildly, and the approximate
equalities~\eqref{eq:roughly_equal_nN} do not hold.

@@

% The projection~\eqref{eq:projection_P} and pull-back~\eqref{eq:pullback_P}
% have the property
% \begin{equation}
%   \label{eq:expectation_pullback}
%   \expeb{f(\yxs) \|\yH} = \expeb{f^*(\yXf)\|\yH}
%   \qquad\text{or}\qquad
%   \sum_{n\yrs=0}^n f(\yrs)\,p_*(\yxs=\yrs) = \sum_{N\yRf=0}^N f^*(\yRf)\,p(\yXf=\yRf).
% \end{equation}
For averaged products, $\sav{\underbrace{\yx\dots\yx}_{\text{$m$
      factors}}}\equiv\binom{n\yxs}{m}/\binom{n}{m}$,
formulae~\eqref{eq:pullback_P} yield the useful equality
\begin{equation}
  \label{eq:expe_products}
%  \label{eq:pullback_m_expectations}
%  \ypp^*\colon \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}
%  \mapsto
    % \bigl( \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}\bigr)^* =
    % \av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}},
    % \\[2\jot]
    % \expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH} =
    % \expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
  \expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH}
\equiv
    % \\[2\jot]
    % {}
    \binom{n}{m}^{-1}
    % \frac{(n-m)!}{n!}
    \sum_{\mathclap{n \yrs=0}}^n %m!\,
    \binom{n \yrs}{m}\, \p(\yxs=\yrs \| \yH)
    % \sum_{N \yRf=0}^N %m!\,
    % \binom{N-m}{N \yRf-m} \binom{N}{N \yRf}^{-1} \p(\yXf=\yRf \| \yH)
    = \binom{N}{m}^{-1}
    % \frac{(N-m)!}{N!}
    \sum_{\mathclap{N \yRf=0}}^N %m!\,
    \binom{N \yRf}{m}\, \p(\yXf=\yRf \| \yH)
\equiv    \expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}.
\end{equation}
The proof of the middle equality uses the expression for the $m$th
factorial moment of the hypergeometric distribution \cite{potts1953}. Thus,
in the state of knowledge $\yH$ the expectations of the averages of
activity products \emph{are initially the same for the sample and for the
  full population}. Similar relations can be found for the raw moments
$\expe{\yxs^m}$ and $\expe{\yXf^m}$, which can be written in terms of the
product expectations via \eqn~\eqref{eq:products_intermsof_average}.
